---
title: "Ouija: Incorporating prior knowledge in single-cell trajectory learning using Bayesian nonlinear factor analysis"
author: "Kieran Campbell"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ggplot2)
theme_set(theme_bw())

set.seed(1L)

knitr::opts_chunk$set(echo = TRUE, # cache = TRUE, 
                      message = FALSE, warning = FALSE,
                      fig.center = TRUE, fig.width = 6, fig.height = 4)
```

## Introduction

`Ouija` allows the incorporation of prior biological knowledge into single-cell trajectory learning in a principled, probabilstic manner. This is ideal for the case when the behaviour of a small number of marker genes for a particular pathway is known a-priori and a trajectory for that pathway needs found.

The input to Ouija should be a cell-by-gene expression matrix of non-negative values. We recommend using `log2(TPM + 1)` or `log2(RPKM + 1)` as this is what the mean-variance relationship in the model is designed for.

Under the hood, Ouija uses Bayesian non-linear factor analysis with priors on the factor loading matrix to specify gene behaviour. Inference is performed using the [Stan](http://mc-stan.org) probabilistic programming language.


## A basic example using synthetic data

Here we can use some synthetic data bundled with the package. This contains a gene expression matrix `synth_gex` comprising 6 genes and 100 cells, along with the true pseudotimes `true_pst`:

```{r load-synth-data}
library(ouija)
data(synth_gex, true_pst)
```

We can quickly plot the gene expression against known pseudotime:

```{r plot-gex, fig.width = 7, fig.height = 3}
dex <- data.frame(synth_gex, true_pst)
names(dex) <- c(paste0("gene", seq_len(ncol(synth_gex))), "pseudotime")
dex_melted <- reshape2::melt(dex, id.vars = "pseudotime", 
                             variable.name = "gene", 
                             value.name = "expression")
ggplot(dex_melted, aes(x = pseudotime, y = expression)) + geom_point() +
  stat_smooth(color = 'red') + facet_wrap(~ gene, scales = "free_y")
```

In order to fit the Bayesian non-linear factor analysis model, we simply call `ouija`. The 
```{r ouija-fit}
oui <- ouija(synth_gex, strengths = 5 * c(1, -1, 1, -1, -1, -1))
print(oui)
```

It's good practice to look at the trace and aurocorrelation of the (log)-likelihood to make sure the distribution has (roughly) converged. More advanced diagnostics may be accessed through the `rstan` package applied to `oui$fit`.

```{r plot-diagnostics}
plot(oui, what = "diagnostic")
```

We can plot the gene expression against the MAP pseudotime too:

```{r plot-map-pseudotime}
plot(oui, what = "behaviour")
```

and heatmaps of gene expression ordered by pseudotime trace:

```{r plot-pseudotime-trace, fig.width = 8}
plot(oui, what = "heatmap", nrow = 3)
```

the MAP pseudotime against the 'true' pseudotime:

```{r plot-map-against-true, fig.width = 4, fig.height = 3}
tmap <- map_pseudotime(oui)
ggplot(data.frame(true_pst = true_pst, tmap = tmap)) + 
  geom_point(aes(x = true_pst, y = tmap)) + 
  xlab("True pseudotime") + ylab("MAP pseudotime") + 
  stat_smooth(aes(x = true_pst, y = tmap), method = "lm", colour = "red") +
  geom_label(x = 0.2, y = 0.8, label = paste("Correlation:", 
                                             format(round(cor(true_pst, tmap), 2), nsmall = 2)))
```



## Inference types

[Stan](http://mc-stan.org) now supports two types of inference: 

* Hamiltonian Monte Carlo (HMC) - full MCMC inference where gradient information of the log-posterior is used to "guide" the random walk through the parameter space
* Automatic Differentiation Variational Bayes (ADVI or simply VI) - approximate inference where the KL divergence to an approximate distribution is minimised

In general, HMC will provide more accurate inference with approximately correct posterior variance for all parameters. However, VB is orders of magnitude quicker than HMC and while it may underestimate posterior variance, anecdotally it seems just as good as HMC for discovering posterior pseudotimes. *If you are using Ouija on more than just a small panel of marker genes it is highly recommended to use Variational Bayes*.

These inference types may be invoked using the `inference_type` argument:

```{r inference-type, results = "hide"}
oui_vb <- ouija(synth_gex, strengths = 5 * c(1, -1, 1, -1, -1, -1), 
                inference_type = "vb")

```

and we can examine the concordance with the HMC result:

```{r compare_hmc, fig.width = 4, fig.height = 3}
ggplot(data.frame(hmc = map_pseudotime(oui), vb = map_pseudotime(oui_vb)),
       aes(x = hmc, y = vb)) +
  geom_point() + xlab("HMC pseudotime") + ylab("Variational Bayes pseudotime")
```


## Technical info

```{r sess-info}
sessionInfo()
```
